---
title: "Seasonal Time Series Analysis"
author: 'Aston Glen Noronha (CWID: 20012232)'
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
<span style="font-size: 18px;font-weight: bold;">Introduction and Motivation </span>

<span style="font-size: 18px;">A seasonal time series of Hotel Hospitality Industry Employees. This dataset contains the number of employees in thousands of persons as monthly averages from 1990-01-01 to 2018-12-01.</span>

<span style="font-size: 18px;">The goal of this project is to predict and forecast the monthly number of employees. </span>

<span style="font-size: 18px;font-weight: bold;">Data Description </span>

<span style="font-size: 18px;">Data Range : From January 1990 to December 2018 </span>

<span style="font-size: 18px;">Dataset Description : The dataset contains number of employee details </span>

<span style="font-size: 18px;">Datasource Description : The data is from Kaggle website and can be accessed using the link: https://www.kaggle.com/datasets/tanmayitanusri/hotel-employees-seasonal-dataset</span>

<span style="font-size: 18px;">Date format : yyyy-mm-dd </span>

<span style="font-size: 18px;">Date : Monthly date </span>

<span style="font-size: 18px;">Employee : number of employee that particular month </span>

```{r}
library(tseries)
library(tidyverse)
library(ggplot2)
library(stats)
library(TSA)
library(forecast)
```
<span style="font-size: 18px;font-weight: bold;">Let's take a look at the dataframe we are working with</span>
```{r}
df = read_csv("C:/Users/Aston/Downloads/Academic Projects/Time Series Project/HotelEmployees.csv")
df <- df %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))
head(df)
```
```{r}
tail(df)
```
<span style="font-size: 18px;font-weight: bold;">Initial look at the data:</span>
```{r}
plot(df, main='Original Data', type='l')
```
```{r}
data = df$Employees
data_12 = ts(data, frequency = 12)
decompose_data = decompose(data_12, 'multiplicative')
plot(decompose_data, type='l', lwd=1, col='blue')
```
```{r}
num_rows <- nrow(df)
train_data <- df[1:(num_rows-12),]
test_data <- tail(df, 12)
data = train_data$Employees
acf(data, lag.max = 200, main='ACF of Original Plot')
```

<span style="font-size: 18px;font-weight: bold;">Doesn't seem to cut off anywhere</span>
```{r}
pacf(data, lag.max = 100, main='PACF of Original Plot')
```

<span style="font-size: 18px;font-weight: bold;">Spikes at regular intervals</span>


<span style="font-size: 18px;font-weight: bold;">Checking if the data is stationary</span>

```{r}
#Checking if the data is stationary
adf_test = adf.test(data)
print(adf_test)
```

<span style="font-size: 18px;font-weight: bold;">Since p-value=0.9769 > 0.05, the TS is non-stationary</span>

```{r}
employ_diff = diff(data)
employ_diff <- ts(employ_diff, frequency=12)
plot(employ_diff, type='l', xlab='Date', ylab='Value', main='Line graph for first-differencing of Employee count')
```
```{r}
adf_test = adf.test(employ_diff)
print(adf_test)
```

<span style="font-size: 18px;font-weight: bold;">Since p-value=0.01 < 0.05, the TS is stationary

```{r}
acf(employ_diff, lag.max = 100, main='ACF of Original Plot')
```
```{r}
pacf(employ_diff, lag.max = 100, main='PACF of Original Plot')
```


<span style="font-size: 18px;font-weight: bold;">There are three significant lags at 12, 24, and 36 which tells us about the order of the seasonal part for the SARIMA model.</span>

<span style="font-size: 18px;font-weight: bold;">The significant lags preceding lag 12 tells us about the order of the non-seasonal part for the SARIMA model.</span>


```{r}
sarima110100 <- arima(employ_diff, order = c(1,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima110200 <- arima(employ_diff, order = c(1,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima110300 <- arima(employ_diff, order = c(1,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima210100 <- arima(employ_diff, order = c(2,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima210200 <- arima(employ_diff, order = c(2,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima210300 <- arima(employ_diff, order = c(2,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima310100 <- arima(employ_diff, order = c(3,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima310200 <- arima(employ_diff, order = c(3,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima310300 <- arima(employ_diff, order = c(3,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima410100 <- arima(employ_diff, order = c(4,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima410200 <- arima(employ_diff, order = c(4,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima410300 <- arima(employ_diff, order = c(4,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima510100 <- arima(employ_diff, order = c(5,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima510200 <- arima(employ_diff, order = c(5,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima510300 <- arima(employ_diff, order = c(5,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima610100 <- arima(employ_diff, order = c(6,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima610200 <- arima(employ_diff, order = c(6,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima610300 <- arima(employ_diff, order = c(6,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima710100 <- arima(employ_diff, order = c(7,1,0), seasonal = list(order = c(1,0,0),period=12))
sarima710200 <- arima(employ_diff, order = c(7,1,0), seasonal = list(order = c(2,0,0),period=12))
sarima710300 <- arima(employ_diff, order = c(7,1,0), seasonal = list(order = c(3,0,0),period=12))
sarima110100
sarima110200
sarima110300
sarima210100
sarima210200
sarima210300
sarima310100
sarima310200
sarima310300
sarima410100
sarima410200
sarima410300
sarima510100
sarima510200
sarima510300
sarima610100
sarima610200
sarima610300
sarima710100
sarima710200
sarima710300
```
```{r}
BIC_values <- c(BIC(sarima110100), BIC(sarima110200), BIC(sarima110300),
                BIC(sarima210100), BIC(sarima210200), BIC(sarima210300),
                BIC(sarima310100), BIC(sarima310200), BIC(sarima310300),
                BIC(sarima410100), BIC(sarima410200), BIC(sarima410300),
                BIC(sarima510100), BIC(sarima510200), BIC(sarima510300),
                BIC(sarima610100), BIC(sarima610200), BIC(sarima610300),
                BIC(sarima710100), BIC(sarima710200), BIC(sarima710300))
BIC_values
```

<span style="font-size: 18px;font-weight: bold;">Finding the lowest BIC value</span>

```{r}
min_BIC <- min(BIC_values)
min_BIC
```
```{r}
min_BIC_index <- which.min(BIC_values)
min_BIC_index
```

<span style="font-size: 18px;font-weight: bold;">From the above BIC values, SARIMA(5,1,0)x(3,0,0) has the lowest value</span>

```{r}
sarima_model <- arima(employ_diff, order = c(5, 1, 0), seasonal = list(order = c(3, 0, 0), period = 12))
residuals = residuals(sarima510300)
acf(residuals, main="ACF of residuals")
```
```{r}
pacf(residuals, main="PACF of residuals")
```

<span style="font-size: 18px;font-weight: bold;">As we can see from the ACF and PACF of the residuals, we can assume that our model has been able to capture almost all of the dependencies. To confirm this assumption, we perform the Ljung's Box Test below.</span>

```{r}
tsdiag(sarima510300)
```

<span style="font-size: 18px;font-weight: bold;">From the above plots, we can confirm that our model has successfully captured all of the dependencies from the data. So we can go ahead with the prediction and forecasting</span>

```{r}
predicted_values <- predict(sarima510300, n.ahead = 12)  
print(predicted_values)
```

<span style="font-size: 18px;font-weight: bold;">These values are the predicted values for the differenced series for the next 12 months. Now we forecast it to get the Original series.</span>

```{r}
original_forecasts <- diffinv(predicted_values$pred, lag = 1, differences = 1, xi = 1957)
original_forecasts <- original_forecasts[original_forecasts != 0]
original_forecasts <- original_forecasts[-1]
original_forecasts
```
```{r}
plot(test_data$Date, test_data$Employees, main = 'Original Data vs Forecast for 2018', type = 'l', col = 'blue', xlab = 'Date', ylab = 'Employees')
lines(test_data$Date, original_forecasts, col = 'red', type = 'l')
legend('bottomright', legend = c('Original Data', 'Forecasts'), col = c('blue', 'red'), lty = 1)
```